Probability

### 2016-04-20 prob.tex
Merge Andy's comments by hand to prob.tex...

@@

The mathematical framework above assumes that one has complete
knowledge of the events and their relationships. But even in many
classical situations, the structure of the event space is only
partially known and the precise dependence of two events on each other
cannot be determined with certainty. In the quantum case, this partial
knowledge is compounded by the fact that not all quantum events can be
observed simultaneously.  Indeed, in the quantum world, there are
non-commuting events which cannot even happen simultaneously. To
accommodate these more complex situations, we completely abandon the
sample space~$\Omega$ and define and reason directly about
events. Thus a quantum probability space will consist of just two
components: a set of events $\qevents$ and a probability measure
$\mu : \qevents \rightarrow [0,1]$. We give an example before giving
the formal definition.

Consider the two-qubit Hilbert space with computational basis
$\ket{0}$ and $\ket{1}$ and states:
\[
\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}, \qquad\ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}\,.
\]
The set of events associated with this Hilbert space consists of all
projections including the empty projection $\mathbb{0}$ and the unit
projection $\mathbb{1} = \proj{0}+\proj{1}$:
\[ 
\{ \mathbb{0}, \proj{0}, \proj{1}, \proj{\ps}, \proj{\ms}, \ldots, \mathbb{1}\}
\]
Each event is interpreted as a possible post post-measurement state of
a quantum system as follows: given some arbitrary current quantum
$\ket{\psi}$ to be measured, the event $\proj{0}$ states that the
post-measurement state will be $\ket{0}$; the event $\proj{1}$ states
that the post-measurement state will be $\ket{1}$; the event
$\proj{\ps}$ states that the post-measurement state will be
$\ket{\ps}$; the event $\proj{\ms}$ states that the post-measurement
state will be $\ket{\ms}$; the event $\mathbb{1}$ states that the
post-measurement state will be a linear combination of $\ket{0}$ and
$\ket{1}$; and the event $\mathbb{0}$ states that the post-measurement
state will be the empty state.

Clearly irrespective of the current state $\ket{\psi}$ and
irrespective of the particular experiment, the probability of event
$\mathbb{0}$ will always be 0 (it is an impossible event) and the
probability of event $\mathbb{1}$ will always be 1 (it is a certain
event). The probabilities attached to other events will depend on the
particular state in question. If the state is $\ket{0}$, the
probability of event $\proj{0}$ is 1; the probability of event
$\proj{1}$ is 0; the probability of event $\proj{\ps}$ is
$\frac{1}{2}$; and the probability of event $\proj{\ms}$ is
$\frac{1}{2}$. If the state is $\ket{\ps}$, the probability of both
events $\proj{0}$ and $\proj{1}$ will be $\frac{1}{2}$; the
probability of event $\proj{\ps}$ is 1; and the probability of event
$\proj{\ms}$ is 0.

@@

We now formalize a \emph{quantum probability space} as
follows~\cite{Maassen2010,Swart2013}. We first assume an ambient
Hilbert space~$\Hilb$ and define the set of events $\qevents$ as all
\emph{projections} on~$\Hilb$. Each quantum state $\ket{\psi}$ induces
a probability measure $\mu_\psi : \qevents \rightarrow [0,1]$ on the
space of events defined as follows:
\[
\mu_\psi(A) = \ip{\psi}{A\psi}
\]
Similarly to the classical case, this probability measure must satisfy:
\begin{itemize}
\item $\mu(\mathbb{1})=1$, and 
\item for a collection of pairwise orthogonal $A_i$, we have
  $\mu(\sum_i A_i) = \sum_i \mu(A_i)$.

@@

Several assumptions are woven in the definition of a quantum probability space:
\begin{itemize}
\item the Hilbert space $\Hilb$;
\item the real interval $[0,1]$; 
\item the fact that each state induces a probability measure, i.e., the
Born rule~\cite{Born1984,Mermin2007}; 
\item the fact that every probability measure is induced by a state, i.e.,
Gleason's theorem~\cite{gleason1957,peres1995quantum,Redhead1987-REDINA}. 
\end{itemize}
In the remainder of the paper, we examine each of these assumptions
and consider variations motivated by computation in a world with limited
resources. In particular, we will consider a variant of the Hilbert
space over finite fields~$\mathbb{F}_{p^{2}}^{d}$~\cite{HansonOrtizSabryEtAl2015,DQT2014,geometry2013}.

>>>>

Probability
theory~\cite{inun.425605319950101,GrahamKnuthPatashnik1994,rohatgi2011introduction}
is defined using the notions of a \emph{sample space} $\Omega$, a
space of \emph{events}~$\events$, and a \emph{probability
  measure}~$\pmeas$. In this paper, we will only consider
\emph{finite} sample spaces: we therefore define a sample space
$\Omega$ as an arbitrary non-empty finite set, the space of events
$\events$ as $2^\Omega$, the powerset of $\Omega$, and the \emph{probability
measure} as a function $\pmeas : \events \rightarrow [0,1]$ such that:
\begin{itemize}
\item $\pmeas(\Omega) = 1$, and 
\item for a collection of pairwise disjoint events $E_i$, the probability
measures are additive $\pmeas(\bigcup E_i) = \sum \pmeas(E_i)$. 
\end{itemize}

\emph{Example of a problem on a finite sample space} (Two coin experiment)
Consider an experiment that tosses two coins. We have four possible outcomes that constitute the
  sample space $\Omega = \{ HH, HT, TH, TT \}$. The event that the
  first coin is ``heads'' is $\{ HH, HT \}$; the event that the two
  coins land on opposite sides is $\{ HT, TH \}$; the event that at
  least one coin is tails is $\{ HT, TH, TT\}$. Depending on the
  assumptions regarding the coins, we can define several probability
  measures. Here is a possible one:
\[\begin{array}{c@{\qquad\qquad}c}
\begin{array}{rcl}
\pmeas(\emptyset) &=& 0 \\
\pmeas(\{ HH \}) &=& 1/3 \\
\pmeas(\{ HT \}) &=& 0 \\
\pmeas(\{ TH \}) &=& 2/3 \\
\pmeas(\{ TT \}) &=& 0 \\
\pmeas(\{  HH, HT \}) &=& 1/3 \\
\pmeas(\{  HH, TH \}) &=& 1 \\
\pmeas(\{  HH , TT \}) &=& 1/3 
\end{array} & \begin{array}{rcl}
\pmeas(\{  HT, TH \}) &=& 2/3 \\
\pmeas(\{  HT , TT \}) &=& 0 \\
\pmeas(\{  TH , TT \}) &=& 2/3 \\
\pmeas(\{  HH, HT, TH \}) &=& 1 \\
\pmeas(\{  HH, HT, TT \}) &=& 1/3 \\
\pmeas(\{  HH, TH, TT \}) &=& 1 \\
\pmeas(\{  HT, TH, TT \}) &=& 2/3 \\
\pmeas(\{  HH, HT, TH, TT \}) &=& 1
\end{array}
\end{array}\]
Note that the probability measure for disjoint events such as $\{ HT \}$ and
$\{ TH \}$ do indeed add.

@@

The mathematical framework above assumes that one has complete
knowledge of the events and their relationships. But even in many
classical situations, the structure of the event space is only
partially known and the precise dependence of two events on each other
cannot be determined with certainty. In the quantum case, this partial
knowledge is compounded by the fact that not all quantum events can be
observed simultaneously.  Indeed, in the quantum world, there are
non-commuting events which cannot even happen simultaneously. To
accommodate these more complex situations, we abandon the
sample space~$\Omega$ and define and reason directly about
events. A quantum probability space consist of just two
components: a set of events $\qevents$ and a probability measure
$\mu : \qevents \rightarrow [0,1]$. We give an example before giving
the formal definition.

Consider the two-qubit Hilbert space with computational basis
$\ket{0}$ and $\ket{1}$ and states:
\[
\ket{\ps}=\frac{\ket{0}+\ket{1}}{\sqrt{2}}, \qquad\ket{\ms}=\frac{\ket{0}-\ket{1}}{\sqrt{2}}\,.
\]
The set of events associated with this Hilbert space consists of all
projections including the empty projection $\mathbb{0}$ and the unit
projection $\mathbb{1} = \proj{0}+\proj{1}$:
\[ 
\{ \mathbb{0}, \proj{0}, \proj{1}, \proj{\ps}, \proj{\ms}, \ldots, \mathbb{1}\}
\]
Each event is interpreted as a possible post post-measurement state of
a quantum system as follows: given some arbitrary current quantum state
$\ket{\psi}$ to be measured, the event $\proj{0}$ states that the
post-measurement state will be $\ket{0}$; the event $\proj{1}$ states
that the post-measurement state will be $\ket{1}$; the event
$\proj{\ps}$ states that the post-measurement state will be
$\ket{\ps}$; the event $\proj{\ms}$ states that the post-measurement
state will be $\ket{\ms}$; the event $\mathbb{1}$ states that the
post-measurement state will be a linear combination of $\ket{0}$ and
$\ket{1}$; and the event $\mathbb{0}$ states that the post-measurement
state will be the empty state.

Irrespective of the current state $\ket{\psi}$ and
irrespective of the particular experiment, the probability of event
$\mathbb{0}$ will always be 0 (it is an impossible event) and the
probability of event $\mathbb{1}$ will always be 1 (it is a certain
event). The probabilities attached to other events will depend on the
particular state in question. If the state is $\ket{0}$, the
probability of event $\proj{0}$ is 1; the probability of event
$\proj{1}$ is 0; the probability of event $\proj{\ps}$ is
$\frac{1}{2}$; and the probability of event $\proj{\ms}$ is
$\frac{1}{2}$. If the state is $\ket{\ps}$, the probability of each
event $\proj{0}$ and $\proj{1}$ will be $\frac{1}{2}$; the
probability of event $\proj{\ps}$ is 1; and the probability of event
$\proj{\ms}$ is 0.

@@

We now formalize a \emph{quantum probability space} as
follows~\cite{Maassen2010,Swart2013}. We first assume an ambient
Hilbert space~$\Hilb$ and define the set of events $\qevents$ as all
\emph{projections} on~$\Hilb$. Each quantum state $\ket{\psi}\in\Hilb$ induces
a probability measure $\mu_\psi : \qevents \rightarrow [0,1]$ on the
space of events defined for any event $A\in\qevents$ as follows:
\[
\mu_\psi(A) = \ip{\psi}{A\psi}
\]
Similarly to the classical case, this probability measure must satisfy:
\begin{itemize}
\item $\mu(\mathbb{1})=1$, and 
\item for a collection of pairwise orthogonal $A_i$, we have
  $\mu(\sum_i A_i) = \sum_i \mu(A_i)$.

@@

Several assumptions are woven in the definition of a quantum probability space:
\begin{itemize}
\item the Hilbert space $\Hilb$;
\item the real interval $[0,1]$; 
\item the fact that each state induces a probability measure, i.e., the
Born rule~\cite{Born1984,Mermin2007}; 
\item the fact that every probability measure is induced by a state, i.e.,
Gleason's theorem~\cite{gleason1957,peres1995quantum,Redhead1987-REDINA}. 
\end{itemize}
In the remainder of the paper, we examine each of these assumptions
and consider variations motivated by computation of numerical quantities in a
world with limited resources. In particular, we will consider a variant of the Hilbert
space over finite fields~$\mathbb{F}_{p^{2}}^{d}$~\cite{HansonOrtizSabryEtAl2015,DQT2014,geometry2013}.
